\subsection{Evaluation Methodology}
\label{sec:eval-method}

We evaluate \textsc{taxonomy\_agent} on two benchmarks chosen to stress
complementary capabilities: (i)~\textbf{20~Newsgroups}~\citep{lang1995newsgroups},
a standard topic-discovery testbed with known-good gold structure, and
(ii)~the \textbf{CoT-Pattern} corpus, a novel safety-relevant benchmark of
chain-of-thought failure modes drawn from the 2024--2026 alignment
literature. The first probes whether the orchestrator--judge loop matches
established topic-modeling baselines on familiar territory; the second
asks whether it can rediscover conceptually subtle patterns whose
linguistic signatures were not designed to be lexically obvious.

\paragraph{Methods.}
We compare four systems on both benchmarks:
(1)~\textsc{taxonomy\_agent} (ours; Claude Sonnet 4.6 orchestrator with
a DeepSeek-v4-Flash judge --- the latter overrides the
installed default of Llama-3.3-70B for a comparable-quality but cheaper
publication pair; see Table~\ref{tab:models});
(2)~\textbf{BERTopic}~\citep{grootendorst2022bertopic}, using
\texttt{all-MiniLM-L6-v2} sentence
embeddings~\citep{reimers2019sbert},
\texttt{UMAP}~\citep{mcinnes2018umap} dimensionality reduction, and
\texttt{HDBSCAN}~\citep{mcinnes2017hdbscan} clustering at library
defaults, with a \texttt{min\_topic\_size}~$\in\{2,5,10,15,25\}$ sweep
reported in \S\ref{sec:eval-results-20ng};
(3)~a \textbf{single-shot LLM} baseline that asks one DeepSeek-v4-Flash
call to propose a taxonomy from a corpus sample, then labels the full
corpus in parallel;
(4)~a \textbf{TopicGPT-style}~\citep{pham2024topicgpt} iterative LLM
proposer with no separate judge. We use ``TopicGPT-style'' rather
than ``TopicGPT'' because this is our reimplementation; relative to
the released \texttt{topicgpt} package
(\url{github.com/chtmp223/topicgpt}) we keep the proposer-refine loop
and the topic-name-plus-definition output format, and drop the
explicit merge-and-prune pass and the human-readable hierarchy
construction step. We did not run the released package; the relative
ordering is a comparison against our reimplementation, not against the
original system.

\paragraph{20~Newsgroups protocol.}
We sample 25 items per class stratified across the 20 classes (a
target of 500), keep the 487 that survive sklearn's default
header/footer/quote stripping (13 documents are emptied by the
filter), repeat under 3 random seeds, and run all 4 methods on each
seed (12 runs total). Gold labels are the 20~Newsgroups class
assignments. We report \emph{external} metrics --- purity, normalized
mutual information (NMI), and adjusted Rand index (ARI) --- against
gold, alongside \emph{intrinsic} topic-quality metrics --- NPMI and
$C_v$ coherence~\citep{rehurek2010gensim} computed from the discovered
cluster--label assignments, plus a redundancy score that penalizes
near-duplicate category labels.

\paragraph{CoT-Pattern corpus.}
We construct a 149-item corpus of chain-of-thought traces exhibiting
five distinct failure modes anchored in the recent safety literature:

\begin{itemize}
  \item \textbf{sycophantic\_capitulation} --- the model abandons a
        correct answer after a user expresses disagreement
        \citep{dubois2026sycophancy};
  \item \textbf{post\_hoc\_rationalization} --- the visible reasoning
        is constructed to justify an answer already chosen
        \citep{liu2026posthoc};
  \item \textbf{unfaithful\_paraphrase} --- the model silently restates
        the problem in a non-equivalent form and solves the new one
        \citep{lanham2023faithfulness};
  \item \textbf{reward\_hack\_verbalization} --- the trace explicitly
        reasons about exploiting graders, test cases, or verifiers
        \citep{hase2026rewardhack};
  \item \textbf{hallucinated\_premise} --- the model introduces a
        non-existent fact or constraint and reasons from it
        \citep{niimi2026hallucination}.
\end{itemize}

Because corpus quality (not orchestrator stochasticity) is the unit of
analysis here, we run each method under a single seed. We acknowledge
this asymmetry with the 3-seed 20NG protocol below; a multi-seed CoT
re-run would tighten confidence intervals but was beyond our compute
budget at the time of writing.
\todo{multi-seed CoT re-run (3 seeds) to bootstrap a confidence
interval around the 0.866 purity number.}

\paragraph{Why a single-source synthetic corpus.}
Before settling on the synthetic design, we surveyed five public CoT
sources and found that none of them, individually or in combination,
supports clean pattern-discovery evaluation across our five patterns.
The Sharma et al.\ sycophancy
stimuli~\citep{sharma2023sycophancy}%
\todo{add bibkey sharma2023sycophancy (``Towards Understanding Sycophancy in Language Models'', 2023)}
are user-side prompts rather than annotated CoT pivots;
the Perez et al.\ model-written
evaluations~\citep{perez2022discovering}%
\todo{add bibkey perez2022discovering (``Discovering Language Model Behaviors with Model-Written Evaluations'', 2022)}
are statistical proxies rather than per-trace structural labels;
FaithCoT-Bench~\citep{paul2024faithcot} conflates arithmetic errors
with hallucinations and does not separate post-hoc rationalisation
from unfaithful paraphrase; the Denison sycophancy-to-subterfuge
transcripts~\citep{denison2024subterfuge} contain only roughly seven
explicit reward-hack-verbalisation examples;
PRM800K~\citep{lightman2023prm800k} labels per-step correctness, not
the named pattern. \textbf{No public dataset provides annotated CoT
pivots for more than one of our five patterns}, and combining
datasets to cover all five makes source detection a near-perfect
proxy for pattern detection (the hybrid corpus, Appendix~A: 95\%
surface-feature accuracy vs.\ 73\% on the single-source corpus,
i.e.\ the hybrid is unevaluable, not merely noisy). Construct-validity
arguments for single-source-synthetic CoT evaluation in this regime
follow the precedents established by~\citet{lanham2023faithfulness}
and~\citet{turpin2023biased}%
\todo{add bibkey turpin2023biased (``Language Models Don't Always Say What They Think'', NeurIPS 2023)}.

\emph{All} 149 items are accordingly drawn from
\textbf{MATH-500 Level~4--5}
problems~\citep{hendrycks2021math} and generated by
\textbf{Gemini-3.1-Flash-Lite} under five pattern-specific injector
prompts. Each generated trace is verified by a
\textbf{Claude Opus 4.8} majority-vote ensemble of three independent
classifiers; only items the ensemble unanimously assigns to the
intended pattern enter the corpus. We note that this verification
ensemble is a different model family from the orchestrator
(Claude~Sonnet), so vendor cross-contamination between
gold-label construction and orchestrator inference is partial rather
than absent --- we flag this in \S\ref{sec:limitations}.

The single-source synthetic corpus is methodologically cleaner than
the hybrid alternative: TF-IDF accuracy drops to 26\% (barely above
the 20\% random baseline), and a 14-feature surface-feature
classifier (length, count of \texttt{\textbackslash boxed},
\texttt{Step}, \texttt{\$}, \ldots), trained under 5-fold
cross-validation on the same 149 items the discovery systems see,
achieves 73\% accuracy. The residual surface signal is structural
rather than lexical --- sycophantic traces are longer, reward-hack
traces mention grading --- and reflects genuine pattern shape rather
than a source artifact.

\paragraph{Interpreting CoT results.}
The 73\% surface baseline is a 5-fold-cross-validated supervised
classifier (logistic regression on the 14 surface features) trained
on the same 149 items the discovery systems see; it is therefore a
\emph{supervised oracle ceiling} for what surface shape alone can do
on this corpus. Taxonomy\_agent's purity, by contrast, is the
optimal-mapping purity of an unsupervised discovered partition.
These two quantities are not strictly commensurable. We therefore
report taxonomy\_agent's CoT result two ways in
\S\ref{sec:eval-results-cot}: (a)~accuracy on the 109 non-\texttt{other}
items, which is the directly comparable quantity to the 73\%
supervised baseline, and (b)~coverage-adjusted purity over all 149
items, treating \texttt{other} as a sixth class. Even with this
clarification, the metrics cannot cleanly distinguish ``modest
semantic discovery on top of surface shape'' from ``sharper surface
extraction with no additional semantic content''
(see \S\ref{sec:limitations}). The random baseline is 20\% (five
balanced classes).

\paragraph{Operational metrics.}
For both benchmarks we additionally report dollar cost (from OpenRouter
\texttt{usage.cost}, broken down by orchestrator and judge) and
wall-clock time. On 20NG we additionally cross-check the
orchestrator's self-reported convergence signal (the judge's
don't-fit rate logged on every classify probe; 12 events total across
the three taxonomy\_agent runs) against terminal external purity, NMI,
and NPMI. The convergence-signal data, Spearman correlations, and the
interpretation that prompts the reframing of the introduction's
``cross-checked'' claim are reported in
\S\ref{sec:eval-results-20ng}; per-iteration partial-purity was not
logged, so the cross-check is conducted at the terminal-pair level
($n{=}3$ seeds).
