\section{Evaluation}
\label{sec:eval}

\input{sections/eval_methodology}

\subsection{Results overview}

We evaluate \textsc{taxonomy\_agent} against three baselines on the
two benchmarks described in Section~\ref{sec:eval-method}, measuring
both intrinsic topic-coherence metrics and external clustering metrics,
alongside cost and wall-time. The evaluation is also designed to
externally validate the system's self-reported convergence signal.
\todo{All numbers pending eval runs.}

\begin{itemize}
  \item \textbf{Corpus.} 20~Newsgroups subset, $K$ seeded categories.
        \todo{Specify $K$, sampling protocol, train/test split.}
  \item \textbf{Methods.} \textsc{taxonomy\_agent} (ours);
        BERTopic~\citep{grootendorst2022bertopic};
        TopicGPT-style prompt~\citep{pham2024topicgpt}; a single-shot
        LLM baseline that asks the orchestrator model for a taxonomy
        in one call.
  \item \textbf{Metrics.} External: purity, NMI, ARI against the
        20~Newsgroups labels. Intrinsic: NPMI and
        $C_v$~\citep{rehurek2010gensim} coherence, label redundancy.
        Operational: dollar cost (from OpenRouter
        \texttt{usage.cost}), wall-clock time, number of iterations.
  \item \textbf{Convergence validation.} Scatter plot of the judge's
        terminal don't-fit rate against external purity / NPMI over
        repeated runs, to test whether the self-reported signal
        correlates with real quality.
  \item \textbf{Tables and figures.}
        \todo{Main results table (methods $\times$ metrics).}
        \todo{Scatter plot of don't-fit rate vs.\ external purity.}
        \todo{Ablation: revise-op vocabulary, judge model, min\_iter
        floor.}
\end{itemize}
