\subsection{20~Newsgroups Results}
\label{sec:eval-results-20ng}

We report mean $\pm$ standard deviation over three seeds
(42, 43, 44) on a 487-document stratified subsample of
20~Newsgroups. All four methods see the same documents
under the same seed; the orchestrator and judge use
\texttt{deepseek-v4-flash} via OpenRouter. Table~\ref{tab:20ng}
gives the headline numbers.

\begin{table*}[t]
\centering
\small
\setlength{\tabcolsep}{4pt}
\begin{tabular}{lcccccccrr}
\toprule
Method & Purity $\uparrow$ & NMI $\uparrow$ & ARI $\uparrow$
       & NPMI $\uparrow$ & $C_v$ $\uparrow$ & Redund.\ $\downarrow$
       & $|\mathcal{T}|$ & Cost (USD) & Wall (s) \\
\midrule
taxonomy\_agent (ours)
  & \textbf{0.367}$_{\pm.020}$ & \textbf{0.656}$_{\pm.012}$ & \textbf{0.336}$_{\pm.035}$
  & 0.065$_{\pm.020}$ & 0.322$_{\pm.021}$ & 0.133$_{\pm.037}$
  & 7.3$_{\pm0.6}$ & 1.86$_{\pm.04}$ & 258$_{\pm57}$ \\
single\_shot
  & 0.253$_{\pm.026}$ & 0.345$_{\pm.056}$ & 0.101$_{\pm.041}$
  & 0.052$_{\pm.016}$ & 0.330$_{\pm.027}$ & 0.144$_{\pm.019}$
  & 10.0$_{\pm0.0}$ & 0.024$_{\pm.000}$ & 169$_{\pm25}$ \\
topicgpt\_style
  & 0.185$_{\pm.094}$ & 0.222$_{\pm.128}$ & 0.043$_{\pm.043}$
  & \textbf{0.106}$_{\pm.041}$ & 0.322$_{\pm.035}$ & \textbf{0.053}$_{\pm.010}$
  & 16.0$_{\pm5.2}$ & 0.030$_{\pm.012}$ & 162$_{\pm60}$ \\
bertopic
  & 0.097$_{\pm.000}$ & 0.159$_{\pm.000}$ & 0.017$_{\pm.000}$
  & 0.007$_{\pm.000}$ & \textbf{0.767}$_{\pm.000}$ & 0.250$_{\pm.000}$
  & 2.0$_{\pm0.0}$ & 0.00 & 15$_{\pm5}$ \\
\bottomrule
\end{tabular}
\caption{20~Newsgroups results (mean $\pm$ stddev over seeds 42--44, $n=487$).
Best per column in bold. $|\mathcal{T}|$ is the number of discovered
categories. BERTopic's $C_v$ leadership is qualified in the text.}
\label{tab:20ng}
\end{table*}

\paragraph{Supervised metrics.}
Against the 20~Newsgroups gold labels, taxonomy\_agent dominates
on all three label-aware metrics. Its NMI of 0.656 is roughly
$1.9\times$ the next-best system (single\_shot at 0.345); its
ARI of 0.336 is $3.3\times$ single\_shot's 0.101; purity is
$1.45\times$ single\_shot's. The margins are consistent across
seeds: taxonomy\_agent's NMI standard deviation is 0.012,
about $4\times$ tighter than single\_shot's 0.056, suggesting
that the agentic loop is converging to a similar
taxonomy structure on each run rather than depending on
prompt luck.

\paragraph{Intrinsic metrics tell a different story.}
Coherence metrics are not in taxonomy\_agent's favour.
topicgpt\_style wins NPMI (0.106 vs.\ our 0.065) by emitting
finer-grained, more lexically-focused categories (16 vs.\ 7);
this also yields the lowest label redundancy. BERTopic wins
$C_v$ by a wide margin (0.767 vs.\ our 0.322), but this number
is misleading: BERTopic collapses the corpus to two macro-clusters
across all seeds at default HDBSCAN parameters
(\texttt{min\_cluster\_size=10}, the library default; we did not
tune it down to force more clusters), so its $C_v$ is the coherence
of two very large, internally diverse topics. A two-cluster solution
can trivially attain high co-occurrence statistics within its top-$n$
words while still failing the supervised metrics catastrophically
(NMI 0.159, purity 0.097). We read this as evidence that intrinsic
coherence and label-aware structure can diverge sharply at this
corpus size, and that practitioners should not optimise $C_v$ alone.

\paragraph{Cost.}
Taxonomy\_agent costs USD\,1.86 per run --- about $77\times$
single\_shot and $62\times$ topicgpt\_style, but in absolute terms
still under USD\,2 for a 487-document corpus. Normalised against the
$+0.31$ NMI gain over single\_shot, this is roughly USD\,5.95 per
NMI-point of improvement at this corpus size; whether that is a
reasonable rate depends on the downstream use case (a one-off
codebook is cheap, a high-frequency drift monitor is not). The cost
is dominated by judge calls during the per-item classification step;
the orchestrator's discovery loop accounts for under 10\%\ of the
spend. Wall-clock time (258\,s on average) is competitive with the
LLM baselines despite the agentic loop, because the rate-limiting
factor for all three LLM methods is per-item judge classification
rather than discovery.

\subsection{CoT-Pattern Results}
\label{sec:eval-results-cot}

The CoT-pattern corpus is a single-seed evaluation on the
synthetic-only \texttt{math500\_l4\_5} corpus described in
\S\ref{sec:eval}: 149 Gemini-3.1-Flash-Lite reasoning traces
across five injected patterns (sycophantic capitulation,
unfaithful paraphrase, post-hoc rationalization,
hallucinated premise, reward-hack verbalization).
We run a single seed because the corpus is small and the gold
labels are deterministic, but we caution that the variance
estimates from Table~\ref{tab:20ng} do not directly transfer.
Table~\ref{tab:cot} summarises.

\begin{table*}[t]
\centering
\small
\setlength{\tabcolsep}{4pt}
\begin{tabular}{lcccccccrr}
\toprule
Method & Purity $\uparrow$ & NMI $\uparrow$ & ARI $\uparrow$
       & NPMI $\uparrow$ & $C_v$ $\uparrow$ & Redund.\ $\downarrow$
       & $|\mathcal{T}|$ & Cost (USD) & Wall (s) \\
\midrule
taxonomy\_agent (ours)
  & \textbf{0.866} & \textbf{0.762} & \textbf{0.660}
  & \textbf{0.033} & \textbf{0.369} & 0.150
  & 5 & 2.23 & 196 \\
single\_shot & 0.336 & 0.143 & 0.036 & 0.025 & 0.348 & 0.203
  & 10 & 0.009 & 34 \\
topicgpt\_style & 0.322 & 0.075 & 0.033 & 0.020 & 0.337 & \textbf{0.075}
  & 3 & 0.011 & 74 \\
bertopic & 0.201 & 0.000 & $-$0.026 & 0.000 & 0.365 & 0.098
  & 6 & 0.00 & 29 \\
\bottomrule
\end{tabular}
\caption{CoT-pattern corpus, single seed (42), $n=149$.}
\label{tab:cot}
\end{table*}

\paragraph{Headline (with caveats and three baselines).}
Taxonomy\_agent attains 0.866 purity, NMI 0.762, ARI 0.660 on the
CoT-Pattern corpus at 73\% coverage (40 of 149 items routed to
\texttt{other}). We report three commensurable comparators
(Appendix~A\footnote{Appendix~A, ``Source/Surface Leakage
Baselines'', gives all baseline definitions, training protocols, and per-class confusion matrix.}):
\textbf{(i) unsupervised k-means} on the same 14 surface features
($k{=}5$, optimal cluster--class mapping via the Hungarian algorithm,
median over 5 random seeds) achieves 0.497 purity --- this is the
true unsupervised head-to-head; \textbf{(ii) supervised
surface-feature LR} (5-fold stratified CV with access to gold) attains
0.73 accuracy as an oracle ceiling; \textbf{(iii) supervised TF-IDF
LR} attains 0.26, barely above the 0.20 random floor, showing the
text-bag is uninformative.

Against the proper unsupervised k-means baseline, taxonomy\_agent's
0.866 is \emph{+37 points}; against the supervised oracle, +14
points. Because the system pushes 27\% of items to \texttt{other},
we also report two coverage-explicit views: \emph{(a)~accuracy on
the 109 items it chose to label} is $104/109{=}95.4\%$, and
\emph{(b)~coverage-adjusted purity over all 149 items} treating
\texttt{other} as a non-match is $104/149{=}0.698$. The honest claim
is that taxonomy\_agent is \emph{high-precision at moderate coverage}:
above all surface baselines on the items it labels, slightly below
the supervised oracle when forced to cover the full corpus.

\paragraph{Interpret with care.}
Three caveats are needed before claiming semantic discovery.
\emph{First}, the corpus is single-domain (competition math) and
synthetic: every CoT was produced by a generator instructed to
inject a specific named pattern, so the patterns are caricatures
rather than naturalistic occurrences. We expect both
taxonomy\_agent and the surface baseline to do worse on
naturalistic data, but it is unclear which would degrade
faster. \emph{Second}, even reading 0.866 as a
precision-at-coverage number, the supervised surface baseline of
0.73 already captures most of the gross structural differences
between patterns (sycophantic CoTs end with capitulation phrases;
reward-hack CoTs name the grader; unfaithful-paraphrase CoTs
mention rewrites). A system that picks up these surface cues plus
a small amount of additional semantic signal would land in roughly
the 0.85--0.95 precision-at-coverage band that we observe on the
109 non-\texttt{other} items, so the gap is consistent with both
``modest semantic discovery on top of surface shape'' and
``exclusively sharper surface-shape extraction''. We cannot
distinguish these from the metrics alone. \emph{Third}, the 27\%
\texttt{other} rate is not uniform across gold patterns: the
hallucinated-premise class is severely under-recovered, with only
9 items assigned vs.\ $\sim$30 in the gold --- the pattern with the
weakest surface tells (no capitulation phrase, no grader mention,
no rewrite marker). Hallucinated-premise items disproportionately
fall into \texttt{other}, which is the strongest single piece of
evidence that the system is leaning on structural shape rather
than pure semantic discovery.

\paragraph{Baselines collapse.}
The contrast with the LLM baselines is sharper here than on
20~Newsgroups: single\_shot reaches only 0.143 NMI and 0.036
ARI, and topicgpt\_style is worse (NMI 0.075). BERTopic's NMI is
indistinguishable from zero. All three LLM methods used the same
DeepSeek model with comparable system prompts and were given the
same probe size and classification instructions; we did not
hand-tune prompts for the baselines. The CoT-pattern task appears
to genuinely require iterative refinement: discovering ``the model
sometimes commits to a wrong answer and rationalises backward''
from a single prompt over a small sample is hard, and the LLM
baselines tend to produce surface-style categories
(``problem-solving traces'', ``mathematical reasoning'') that
collapse the pattern distinction.

\subsection{Qualitative Analysis}
\label{sec:eval-results-qual}

We give a small qualitative window into the discovered taxonomies.

\paragraph{20~Newsgroups (seed 42).} Taxonomy\_agent discovered
8 categories. Three representative ones, with our best mapping to
the 20-newsgroup gold labels:
\begin{itemize}
  \item \emph{computers\_and\_software} (``operating systems, software,
    programming, Windows, DOS, general computing'') --- recovers
    \texttt{comp.os.ms-windows.misc} and \texttt{comp.sys.ibm.pc.\linebreak hardware} cleanly.
  \item \emph{politics\_and\_society} (``politics, social issues, crime,
    punishment, gun control, Middle East politics, civil rights'') ---
    correctly fuses \texttt{talk.politics.misc}, \texttt{talk.politics.guns},
    and \texttt{talk.politics.mideast}, which is arguably the right move
    semantically but counts against ARI because the gold taxonomy keeps them
    separate.
  \item \emph{recreation\_and\_sports} (``recreational activities, sports,
    motorcycles, hobbies, for-sale listings'') --- pools four gold
    classes (\texttt{rec.autos}, \texttt{rec.motorcycles},
    \texttt{rec.sport.baseball}, \texttt{rec.sport.hockey}) and bleeds in
    \texttt{misc.forsale}. The fusion is defensible but again costs
    label-aware credit.
\end{itemize}
This pattern --- semantically coherent macro-clusters that
nonetheless lose ARI to the fine-grained 20-way gold partition ---
is consistent across seeds 43 and 44, which produced 7 categories each.
We read it as evidence that the system's failure mode on supervised
metrics is \emph{lumping}, not \emph{splitting}: it converges on a
mid-granularity taxonomy that a human evaluator would call reasonable
and that nevertheless costs supervised credit on a corpus whose
gold labels are very fine-grained.

\paragraph{CoT-pattern (seed 42).} The agent discovered exactly five
categories, and their names match the five gold patterns
(\emph{sycophantic\_capitulation}, \emph{unfaithful\_paraphrase},
\emph{post\_hoc\_rationalization}, \emph{hallucinated\_premise},
\emph{reward\_hack\_verbalization}). This is suspicious in two
directions. \emph{First}, the orchestrator was told the
classification instruction (``categories should reflect distinct ways
the model is reasoning''), but not the pattern names: name agreement
is therefore non-trivial and is at least partial evidence of
semantic discovery rather than surface clustering. \emph{Second}, the
agent had access to the corpus only through its tools and a small
prompted instruction; the orchestrator did not see any pattern labels.
That said, the agent's class-count distribution (sycophantic 26,
post-hoc 19, paraphrase 28, hallucinated-premise 9, reward-hack 27,
other 40) shows the same under-recovery of the
\emph{hallucinated\_premise} class noted above:
hallucinated-premise traces lack the strong lexical tells of the other
four patterns (no capitulation phrase, no grader mention,
no rewrite marker), and the agent appears to fall back on
\texttt{other} when those tells are absent. This is the
strongest single piece of evidence that the system's headline
accuracy is partly driven by structural shape rather than purely
by semantic understanding of the constructs.

\paragraph{Summary.} Across both corpora, taxonomy\_agent's
strength is label-aware structure (NMI, ARI, purity), particularly
where the task requires iterating from a partial picture to a
coherent set of mid-granularity categories. Its weaknesses are
intrinsic coherence ($C_v$, where BERTopic's degenerate two-cluster
solution wins), per-call cost (roughly USD\,2 per corpus vs.\
USD\,0.02--0.03 for the LLM baselines), and the same failure mode
in both corpora: when a class lacks strong surface tells
(\texttt{misc.forsale} in 20~Newsgroups, \texttt{hallucinated\_premise}
in CoT), it is absorbed into a sibling category or pushed into
\texttt{other}.
