\section{Introduction}
\label{sec:intro}

Constructing a labelled taxonomy over an unlabelled corpus is a recurring
bottleneck in applied NLP. Annotation guidelines for supervised
classification, codebooks for qualitative research, summarisation schemas
for analyst-facing dashboards, and policy hierarchies for content
moderation all begin life as a manual taxonomy. Producing one typically
requires domain experts to read hundreds of documents, draft candidate
categories, argue over overlaps, and iterate until coverage and
disjointness are acceptable. The process is slow, expensive, and difficult
to reproduce, and the resulting taxonomy is often a brittle artefact that
nobody is incentivised to revisit when the underlying corpus drifts.

Recent work has shown that large language models (LLMs) can replace
much of this manual labour. TopicGPT~\citep{pham2024topicgpt} prompts an
LLM to generate topic names and definitions directly from sampled
documents (including a refinement pass), while
TnT-LLM~\citep{wan2024tntllm} stages a two-phase pipeline that first
summarises a corpus into candidate labels and then distils a
lightweight classifier. Iterative LLM edit loops also appear in
goal-driven and human-in-the-loop systems such as
GoalEx~\citep{wang2023goalex} and LLooM~\citep{lam2024lloom}.
Co-DETECT~\citep{codetect2025} is last year's EMNLP System
Demonstration on collaborative codebook construction with LLMs and
the closest peer to our submission; we sit at an orthogonal operating
point (taxonomy discovery from scratch over an unlabelled corpus,
without a seed sketch and without a human in the loop) and discuss
the relationship in \S\ref{sec:related}. Against this backdrop, our contribution
is a careful engineering recomposition; we make no claim of a wholly new mechanism, but we
narrow the orchestrator's edit surface to a typed validated DSL with
per-operation failure isolation, decouple labelling into a parallel judge,
and ship the result with operational tooling for live cost accounting,
partial-progress saving, and a web UI. These are deliberate practical
choices for an operational demo, and we make no claim of architectural
novelty.

We present \textsc{taxonomy\_agent}, an open-source system that
packages these choices. The system is structured as a two-tier agent: a
capable \emph{orchestrator} LLM holds long-horizon state and proposes
edits via a six-tool interface, while a cheap \emph{judge} LLM runs in
parallel to score items, flag don't-fit cases, and propose novelties.
Edits are constrained to a typed revise DSL with six operations
(add, rename, edit, merge, split, drop), each validated before
mutation, with per-operation failure isolation. Convergence is
self-reported via the judge's don't-fit rate with a minimum iteration
floor and an explicit escape hatch; we cross-check this self-report
against external purity and NMI on 20~Newsgroups, but stop short of
claiming the don't-fit rate is a quantitatively validated leading
indicator.

Empirically, the system outperforms a one-shot LLM baseline on
20~Newsgroups by $1.9\times$ on NMI and $3.3\times$ on ARI, with
wider margins over a minimal iterative LLM baseline, and retains a
$1.3\times$ NMI and $1.4\times$ ARI lead over the strongest
LLM-driven comparator we run, an embed-then-LLM-label pipeline
(\S\ref{sec:eval-results-20ng}). The lead over BERTopic at library
defaults is wide but narrows under a sweep of BERTopic's minimum
cluster size; the ARI lead is robust at $2.6\times$ over the
best-tuned BERTopic on 20NG. The
iterative baseline is our own minimal port of TopicGPT's idea; we did
not run the released TopicGPT package and make no claim about ordering
against TopicGPT as published. On the synthetic CoT-Pattern corpus the system
labels 109 of the 149 items with 100\%\ accuracy and reaches 0.732
purity over the full corpus, with precision on labelled items at
$100\%$ across each of the four mathematical sub-domains
(Table~\ref{tab:cot_per_domain}). Against a supervised ceiling of 0.730 from
a surface-feature classifier with gold-label access, this is $+27$
points on the items the system labels and at parity at full coverage;
against an unsupervised k-means baseline on the same surface features
it is $+23$ points at full coverage. As we discuss
in \S\ref{sec:eval-results-cot}, the metrics alone cannot cleanly
attribute the margin to semantic discovery and rule out sharper surface
extraction.

The output primitive is a partitioning taxonomy with single labels
plus an explicit \texttt{other} escape, distinguishing it from
concept-induction systems such as LLooM~\citep{lam2024lloom} that
surface overlapping high-level axes. What turns these design choices
into a tool a stranger can use are the operational pieces around them:
live native cost accounting broken down by orchestrator and judge,
partial-progress streaming to disk so a long run is interruptible,
vendor-agnostic routing, and a web UI for live inspection of cost,
taxonomy state, and per-item classifications.
Our contributions are:
\begin{itemize}
  \item \textbf{Operational tooling for a tool a stranger can use}
        (\S\ref{sec:eval-demo}). Live native provider cost accounting
        broken down by orchestrator and judge, partial-progress
        streaming to disk so a killed run leaves recoverable state,
        vendor-agnostic routing across any chat-completions model on
        the provider, a web UI that surfaces all of the above live,
        and a disciplined test suite that locks each fixed bug with a
        named regression test. We claim this layer as the strongest
        demonstration contribution.
  \item \textbf{A typed dispatcher that validates before mutating} over
        six standard verbs (\textit{add, rename, edit, merge, split,
        drop}) with per-operation failure isolation, providing an
        auditable trace.jsonl of every edit. A single-seed ablation
        (\S\ref{sec:eval-results-cot}) replaces the typed dispatcher
        with a loose dispatcher that accepts ops verbatim; the
        orchestrator stopped issuing revise calls and we terminated
        the run with an empty taxonomy. The contrast is suggestive
        but a single seed cannot isolate the dispatcher from ordinary
        stochasticity; we report this as a preliminary observation,
        not validated support. We do not claim the verbs themselves
        are novel.
  \item \textbf{A parallel stateless judge architecture}, building on
        the generator and labeller split of
        TnT-LLM~\citep{wan2024tntllm}, and a don't-fit rate that the
        orchestrator treats as a termination signal. We cross-check this
        signal against external clustering metrics in
        \S\ref{sec:eval-results-20ng}; it saturates at the floor and
        does not function as a quantitative leading indicator at the
        sample size available.
  \item \textbf{An open-source artefact at}
        \repourl{} with a
        \texttt{pip install} install path, a 60-second \emph{Quick
        demo} button, and reproducible 20NG and CoT benchmark
        configurations.
\end{itemize}
