\section*{Limitations}
\label{sec:limitations}

\paragraph{The CoT margin is consistent with sharper surface
extraction.} Our primary CoT result, 95.4\%\ accuracy on the 109
items the system labels, against a supervised baseline of 0.730
trained on 14 surface features over the full 149 items, lies
within the range that a sharper surface extractor could plausibly
produce. The supervised metrics alone cannot distinguish (a)~modest
semantic discovery on top of structural pattern shape from
(b)~sharper extraction of that same structural shape with no
additional semantic content. The under-recovery of
\texttt{hallucinated\_premise} (9 of $\sim$30 gold items, the pattern
with the weakest surface tells) is the strongest single piece of
evidence for hypothesis~(b). A convincing test would require either
a corpus where surface and semantic signal are dissociated by
construction, or a naturalistic spot-check with hand-curated real
CoT pivots; we have neither, and we do not claim the CoT result
settles the question.

\paragraph{Self-reported convergence is circular even with a
model split, and saturates at the floor in practice.} The don't-fit
rate that terminates the loop is computed by the same judge LLM that
produced the labels. The orchestrator--judge model split (Sonnet
orchestrator, DeepSeek-v4-Flash judge in the eval; Llama-3.3-70B in
the installed default) breaks the cross-role training-data overlap but
does not break the loop: a stable terminal don't-fit just says the
judge is self-consistent with respect to its own labelling function.
Empirically the signal also saturates at the floor too quickly to
function as a quantitative leading indicator: across the 12 classify
events in the three 20NG runs, 9 of 12 report \texttt{dont\_fit\_rate}
$\leq 0.05$ and 7 of 12 report exactly 0.0, including a
non-monotone $0.0 \to 0.05$ bounce in one of the three seeds
(\S\ref{sec:eval-results-20ng}). In the deployment case the system
is sold for, discovery on a corpus with no gold, external
validation is exactly what is unavailable, and the convergence
signal alone should not be treated as evidence the discovered
taxonomy is correct.

\paragraph{Researcher degrees of freedom in convergence calibration.}
Both the orchestrator prompt and the judge prompt were iterated by
the authors on the 20~Newsgroups and CoT corpora during development.
This means the $0.10$ don't-fit threshold, the
\texttt{min\_iterations} floor, and the verb-level instructions are
calibrated against the same benchmarks we evaluate on, instead of
being discovered from scratch on each new corpus. We did not hold out a
separate calibration corpus. A reviewer should treat the convergence
behaviour as fitted to our test corpora; the convergence properties
do not hold independently of corpus.

\paragraph{Single-seed CoT, no confidence interval.}
The 20NG numbers are reported over three independently sampled random
seeds with standard deviations; the CoT result is single-seed. Even
with deterministic gold labels, the orchestrator and judge are stochastic
LLMs, so the 0.866 number has no confidence interval. The
asymmetry between the two benchmarks is a real limitation; it is not a
principled choice, and a multi-seed CoT re-run is the most
important follow-up experiment.

\paragraph{Potential gold-label vendor contamination.}
The CoT corpus is generated by Gemini under named-pattern injectors
and verified by a Claude~Opus~4.8 majority-vote ensemble. The
orchestrator is Claude~Sonnet, a different model size in the same
family as the verifier. This means the gold labels themselves encode
whatever heuristics the verifier uses, and the orchestrator may
share those heuristics. For 20~Newsgroups, both Claude and the
DeepSeek/Llama judges are almost certain to have been trained on this
public corpus, so the 0.656 NMI may partly reflect memorisation of
the gold partition.

\paragraph{The CoT corpus is synthetic and single-domain.}
Our 149-item CoT benchmark covers competition mathematics only, with
each trace generated by an LLM under a pattern-specific injector
prompt. The synthetic-only choice is, as we argue in
\S\ref{sec:eval-method}, the only methodologically sound option
given the public CoT-pattern data landscape (no public dataset
covers more than one of our five patterns; the hybrid alternative
has 95\% surface leakage). This trade-off does carry a cost: we have
\emph{no evidence} that the system works on naturalistic CoT
failures, and a reviewer who interprets ``synthetic single-source'' as
``the authors couldn't get real data to work'' is not entirely
wrong. A corpus with multiple sources per pattern and explicit
deconfounding is the natural next step, but was beyond the scope of
this paper.

\paragraph{Cost scales with iteration count.}
Orchestrator input tokens grow with conversation history, so a run
that converges in $k$ iterations spends $O(k^2)$ input tokens in the
worst case; at default settings this is roughly \$2 per 100 items.
Anthropic prompt caching via OpenRouter would amortise the repeated
prefix, but we verified empirically that LangChain's
\texttt{bind\_tools} path strips the cache-control markers before the
request leaves the client, so the cache is never hit. We flag a
native-SDK rewrite as future work; we do not fix it in this paper.

\paragraph{Prompt-sensitivity of the judge's escape hatch.}
The don't-fit rate is meaningful only because the classify prompt
includes a fixed \texttt{other}-escape suffix and post-hoc coercion
maps out-of-taxonomy labels to \texttt{other}. Both choices were
prompt-tuned during development; we have not measured how sensitive
the terminal don't-fit rate is to variations in the escape-hatch
wording, and a more permissive or more restrictive escape could
shift the primary numbers materially.

\paragraph{The five-pattern selection is non-exhaustive.}
We chose the five failure modes for their citation centrality in the
2024--2026 CoT-faithfulness literature and for the fact that they can
be elicited in a single trace. Several phenomena of current interest
to the alignment community, including evaluation-awareness,
alignment-faking, and fully internalised (non-verbalised) reasoning,
require multi-turn or counterfactual setups and are out of scope
for a benchmark that scores single chain-of-thought traces.

\paragraph{Language and domain coverage.}
All CoT experiments are English-only and mathematics-only; the
20~Newsgroups runs are English-only news posts. We provide no
evidence about cross-lingual transfer of the typed revise
vocabulary, nor about generalisation to long-form scientific text,
code, multi-turn dialogue, or non-mathematical reasoning domains.

\paragraph{No component-level ablations.}
The four architectural choices we describe in
\S\ref{sec:system}, namely the orchestrator/judge model split, the
typed revise DSL that validates each operation before applying it, the
don't-fit terminal criterion, and the parallel judge dispatcher,
are not independently ablated in
\S\ref{sec:eval-results-20ng}--\S\ref{sec:eval-results-cot}. The
empirical comparison set is taxonomy\_agent end-to-end against three
external baselines, plus the indirect comparison between
\textsc{single\_shot} and \textsc{iterative\_proposal} that contrasts
a one-shot proposal with an iterative one that shares the
orchestrator and judge roles. We are therefore unable to isolate which
architectural choice drives the gains. The most informative single missing experiment is
taxonomy\_agent with free-form prose revise instead of the typed DSL on
the CoT corpus; we estimate roughly USD\,0.30 of OpenRouter spend for
a single-seed run.

\paragraph{Missing direct baseline comparisons.}
Four systems we discuss in~\S\ref{sec:related} are not run as
baselines: \textbf{Co-DETECT} requires a seed sketch codebook and a
human in the loop, which our pipeline does not produce, and no
released artefact ran on either of our benchmarks at submission
time; \textbf{TaxoAdapt} requires a seed taxonomy to adapt, whereas
we discover from scratch; \textbf{GoalEx} ran on a different
benchmark family with metrics we do not collect; \textbf{TnT-LLM} is
an offline production pipeline that distils a separate classifier
and was not packaged for academic comparison. We name these gaps so
a reader can weigh them against our results; the closest comparator
we actually run, the minimal iterative-proposal port, deliberately
lacks TopicGPT's merge-and-prune pass and is not a substitute for a
direct comparison against the released artefact.

\paragraph{Untested vendor-agnosticism and resume from kill.}
The system is engineered for vendor agnosticism via OpenRouter and
for partial-progress streaming to disk, but the operational
properties we report in~\S\ref{sec:eval-demo} are properties of the
code path and the cost tracker; they are not validated quality
properties. We did not run a quality comparison against a
non-Anthropic orchestrator, and re-launching a killed run to continue
from the saved \texttt{taxonomy\_state.json} is not implemented: the
Streamlit UI displays partial state for inspection but does not
resume the orchestrator.
