\section{Related Work}
\label{sec:related}

\paragraph{LLM-based topic discovery.}
TopicGPT~\citep{pham2024topicgpt} prompts an LLM to generate
human-readable topic labels and short definitions directly from sampled
documents, then assigns each remaining document to one of the generated
topics; it includes a refinement pass that merges and prunes topics
discovered in earlier rounds. It demonstrated that LLM-produced topics
are substantially more interpretable than classical BERTopic clusters,
but the refinement is operationally bundled into the prompt instead of
being expressed as a typed edit vocabulary, and the system is not
released as an interactive tool with first-class cost or
partial-progress accounting. TnT-LLM~\citep{wan2024tntllm} from
Microsoft stages a heavier two-phase pipeline in which an LLM first
summarises batches of a conversational corpus into candidate labels,
and a second phase distils a lightweight classifier that scales the
labelling to the full corpus. TnT-LLM is closer in spirit to our
setting, since it already separates a label-generator role from a
labeller role, anticipating our orchestrator--judge split; its
``refine'' step, however, is a single batched summarisation; it is not a
per-item dispatch loop driven by items that do not fit, and it is
structured as an offline production pipeline. Our
\texttt{embed\_cluster\_llm} baseline in Table~\ref{tab:20ng} is a
minimal reimplementation of the TnT-LLM generator-then-labeller
shape (sentence-transformer embeddings, $K{=}16$ k-means, one LLM
naming call per cluster, then LLM per-document assignment) for
direct comparison.

\paragraph{Classical topic modelling.}
BERTopic~\citep{grootendorst2022bertopic} represents the strongest
non-LLM baseline: it embeds documents with a sentence transformer,
reduces dimensionality with UMAP, clusters with HDBSCAN, and labels
clusters via class-based TF-IDF. It is fast, fully reproducible, and
needs no API key, but its outputs are clusters described as bags of keywords, with no
human-readable category names or definitions,
and the taxonomy cannot be edited in a typed way once produced.
Classical LDA~\citep{blei2003lda} remains a reference point but
suffers from the same interpretability and editability issues at a
larger margin.

\paragraph{Co-DETECT.}
Co-DETECT~\citep{codetect2025} is last year's EMNLP System
Demonstration on collaborative codebook construction with LLMs. The
human and the model alternate proposals and revisions: a domain
expert supplies a sketch-level codebook and a small dataset; the LLM
annotates and surfaces edge cases that the sketch fails to cover;
the human accepts, rejects, or amends each edge-case rule, and the
revised codebook is then redeployed. The artefact targets the
ongoing refinement loop a policy team would sit inside for weeks.
\textsc{taxonomy\_agent} sits at an orthogonal operating point: we
discover a candidate taxonomy from scratch over an unlabelled corpus
without a seed sketch and without a human inside the loop, then
hand off the trace.jsonl, the working taxonomy, and the per-item
classifications for downstream human review or for Co-DETECT-style
refinement to take over. We do not run a direct comparison against
Co-DETECT because the input contracts differ (seed codebook plus edge
cases versus unlabelled corpus plus goal instruction), and because no
publicly released Co-DETECT artefact ran on either of our two
benchmarks at the time of submission. For mixed-initiative codebook
construction with a policy expert in the loop, Co-DETECT is the
better fit; \textsc{taxonomy\_agent} is the upstream tool that
produces the candidate codebook Co-DETECT then refines.

\paragraph{Goal-driven and mixed-initiative systems.}
GoalEx~\citep{wang2023goalex} formulates explanation as goal-driven
clustering: a user specifies a natural-language goal and the system
produces a set of explanations partitioning the corpus that satisfy
that goal. We inherit this input contract, since the \texttt{instruction}
and \texttt{category\_focus} parameters in our run configuration
realise exactly GoalEx's goal-conditioned input, and package it
for an interactive demo with a typed revise dispatcher, live cost
accounting, and partial-progress streaming. LLooM~\citep{lam2024lloom}
is a CHI~2024 mixed-initiative system that surfaces high-level
concepts from unstructured text for analysts to inspect and edit
interactively. The LLooM concept-induction primitive is different
from ours: LLooM outputs overlapping high-level axes that a single
document can score on multiple of, whereas \textsc{taxonomy\_agent}
outputs a partitioning taxonomy with single labels plus an
\texttt{other} escape. Choosing between them is a question of
downstream use: a thematic analyst inspecting concepts wants LLooM's
overlapping axes, whereas a downstream classifier or content-policy
hierarchy wants our partition.

\paragraph{Recent 2025 work.}
TaxoAdapt~\citep{taxoadapt2025} adapts an existing taxonomy to a new
domain via LLM-guided edits, sharing our intuition that taxonomy
construction is naturally an edit problem, more closely matched by
iterative revision than by single-shot generation; its edit
vocabulary overlaps with ours
(add/merge/split are natural for taxonomy maintenance), and it
focuses on taxonomy \emph{adaptation} given a seed, whereas we
discover from scratch under a typed dispatcher that validates each
operation before applying it, with isolation between operations on
failure.
LLM-guided semantic-aware clustering~\citep{llmguided2025} uses an LLM
to steer the clustering objective itself, blurring the line between
classical clustering and LLM topic discovery. The recent EMNLP~2024
survey on LLMs for data annotation~\citep{tan2024llmannot} places this
family of work in the broader annotation landscape and motivates the
operational concerns (cost, reproducibility, interruption-safety)
that our system targets.

\paragraph{Positioning of \textsc{taxonomy\_agent}.}
We do not claim a wholly new mechanism: TopicGPT iterates, TnT-LLM
separates a generator from a labeller, ReAct-style
agents~\citep{yao2022react} structure tool-using LLMs, and the
add/rename/merge/split/drop verbs are the standard taxonomy-edit
vocabulary. The contribution is the operational recomposition the
introduction enumerates: first-class operational tooling that prior
LLM topic-discovery and codebook systems treat as out of scope, a
typed dispatcher with per-operation failure isolation, and a parallel
stateless judge with an \texttt{other} escape hatch and a saturated
don't-fit rate. The empirical case is in~\S\ref{sec:evaluation}.
