\section{System Overview}
\label{sec:system}

\textsc{taxonomy\_agent} treats taxonomy discovery as a tool-use
problem: an orchestrator LLM plans and edits, a separate judge LLM
labels, and a typed revise DSL guarantees that every taxonomy mutation
is auditable and atomic. The loop terminates when the judge's
\emph{don't-fit rate} drops below a configurable threshold, gated by a
minimum-iteration floor. This section describes the architecture
(\S\ref{sec:system:arch}), the six tools that form the orchestrator's
action space (\S\ref{sec:system:tools}), the revise DSL
(\S\ref{sec:system:revise}), the convergence criterion
(\S\ref{sec:system:convergence}), and the operational safeguards that
make the system usable as a walk-up demo
(\S\ref{sec:system:safeguards}).

\subsection{Architecture}
\label{sec:system:arch}

Figure~\ref{fig:architecture} sketches the run-time topology.
\begin{figure}[t]
\centering
\begin{tikzpicture}[
  font=\footnotesize,
  node distance=4mm and 6mm,
  box/.style={draw, rounded corners=2pt, align=center, inner sep=3pt,
              minimum height=7mm},
  orch/.style={box, fill=blue!8},
  judge/.style={box, fill=red!8},
  store/.style={box, fill=gray!12},
  ui/.style={box, fill=green!8},
  flow/.style={-{Stealth[length=3pt]}, thick},
  back/.style={-{Stealth[length=3pt]}, dashed, thick},
]
% Corpus on the left
\node[store] (corpus) {Corpus\\(unlabelled)};
% Orchestrator
\node[orch, right=of corpus] (orch) {Orchestrator\\Sonnet 4.6};
% Revise DSL
\node[box, above=of orch, yshift=-2mm] (revise)
  {Typed revise DSL\\\textsc{add} \textsc{rename} \textsc{edit}\\\textsc{merge} \textsc{split} \textsc{drop}};
% Working taxonomy
\node[store, right=of revise] (tax) {Working\\taxonomy};
% Judge pool
\node[judge, below=of orch, yshift=2mm] (judge)
  {Judge pool (parallel)\\DeepSeek-v4-Flash};
% Per-item labels
\node[store, right=of judge] (labels) {Per-item\\labels};
% Cost tracker
\node[store, below=of corpus, yshift=2mm] (cost) {Cost tracker\\\texttt{usage.cost}};
% Partial-progress sink
\node[store, below=of judge, yshift=2mm] (sink)
  {Partial-progress\\sink (disk)};
% Streamlit
\node[ui, right=of sink] (ui) {Streamlit UI};

% Flows
\draw[flow] (corpus) -- (orch);
\draw[flow] (orch) -- node[midway, right, font=\scriptsize]{edit} (revise);
\draw[back] (revise) -- node[midway, left, font=\scriptsize]{validate} (orch);
\draw[flow] (revise) -- (tax);
\draw[flow] (orch.south east) -- node[midway, above, font=\scriptsize]{classify} (judge.north east);
\draw[back] (judge.north west) -- node[midway, below, font=\scriptsize]{don't-fit} (orch.south west);
\draw[flow] (tax.south) |- (judge.east);
\draw[flow] (judge) -- (labels);
\draw[flow] (orch.west) |- (cost.east);
\draw[flow] (judge) -- (sink);
\draw[flow] (sink) -- (ui);
\draw[flow] (cost.south) |- ($(sink.west)+(0,-2mm)$);
\end{tikzpicture}
\caption{Run-time topology. The orchestrator (Sonnet 4.6) holds the
working taxonomy as closure state and edits it through a typed
revise DSL of six verbs; each edit is validated before mutation.
The judge pool (DeepSeek-v4-Flash, run in parallel) labels items
against the current taxonomy and returns a don't-fit rate that
drives termination. A cost tracker and a partial-progress sink
stream to disk; the Streamlit UI reads both for live inspection
during a run and post-hoc browsing afterwards.}
\label{fig:architecture}
\end{figure}
Two roles are kept strictly separate. The \textbf{orchestrator} is a
ReAct-style~\citep{yao2022react}
agent built on LangGraph's
\texttt{create\_react\_agent} \citep{langgraph}; it owns the working
taxonomy as closure state and
emits tool calls to inspect the corpus, edit the taxonomy, or dispatch
work to the judge. The \textbf{judge} is a stateless function that
labels one item at a time against the taxonomy the orchestrator hands
it; it never sees the taxonomy history or the orchestrator's
reasoning. Keeping the two roles separate is intended to have three
effects: (i) a weak orchestrator cannot bluff a label past the judge,
because the judge re-derives it from the item text; (ii) the
orchestrator's context never bloats with per-item labels, so its
prompt stays small across long corpora; and (iii) the judge is
trivially parallelisable via a thread pool. All three effects are
architectural commitments we did not separately ablate; the wall-clock
and cost numbers in \S\ref{sec:eval-results-20ng} are end-to-end and
reflect both the parallel-judge dispatcher and the orchestrator-loop
overhead together.

Both LLMs are called through OpenRouter, so any chat-completions model
can fill either role. The orchestrator defaults to
\texttt{anthropic/claude-sonnet-4.6} (chosen for reliable tool-call
formatting); the judge defaults to
\texttt{meta-llama/llama-3.3-70b-instruct} (chosen for cost). Both can
be swapped at the command line or in the UI without touching code, and
mixed-vendor runs (for instance, a Claude orchestrator paired with a
Llama judge) are the common case in our evaluation
(\S\ref{sec:evaluation}).

Internally, the orchestrator runs as a LangGraph state machine: the
ReAct loop alternates between an \emph{LLM} node that emits the next
tool call and a \emph{tools} node that executes it and appends the
result to the message list. State persists in two places. The
\emph{LangGraph message history} is the orchestrator's working memory
(reset each run). The \emph{closure state} held by the tool factory
(\texttt{\_TaxonomyState} in \texttt{tools.py}) holds the working
taxonomy, the set of item ids already sampled, a snapshot of the
last-finalised taxonomy, and a budget counter for classify calls. The
taxonomy lives only in this closure, so the orchestrator never has to
pass it as an argument, which both shrinks tool-call payloads and
removes a class of bugs where the agent edits a stale copy.

\subsection{The Six Tools}
\label{sec:system:tools}

The orchestrator's entire action space is six tools, summarised in
Table~\ref{tab:tools}. We deliberately kept the surface small: any
additional verb (e.g.\ ``inspect a single category'', ``compare two
items'') can be expressed as a composition of \texttt{sample\_items},
\texttt{get\_taxonomy}, and a freshly-worded
\texttt{classify\_with\_judge} prompt, which the orchestrator
discovers without us having to specify it.

\begin{table*}[t]
\centering
\begin{tabular}{@{}p{0.27\textwidth}p{0.69\textwidth}@{}}
\toprule
\textbf{Tool} & \textbf{Behaviour} \\
\midrule
\texttt{sample\_items(k)} &
Returns $k$ unseen items from the corpus pool. Tracks already-sampled
ids so probes do not repeat until the pool is exhausted, then resets
with a warning to the orchestrator. \\
\texttt{get\_taxonomy()} &
Returns the current working taxonomy as JSON. Read-only. \\
\texttt{revise\_taxonomy(ops)} &
Applies an ordered list of typed operations
(\textsc{add}/\textsc{rename}/\textsc{edit}/\textsc{merge}/\textsc{split}/\textsc{drop})
through the validating dispatcher of \S\ref{sec:system:revise}.
Returns a per-op log and the resulting taxonomy. \\
\texttt{classify\_with\_judge(ids, prompt)} &
Dispatches the named items to the judge in parallel, coerces each
reply to a valid label (or the literal \texttt{"other"}), and returns
per-item labels plus the don't-fit rate. Capped by a per-run budget. \\
\texttt{propose\_novelties\_with\_judge(ids, prompt)} &
Asks the judge to read items that did not fit and suggest new
categories. The judge sees the current taxonomy so its proposals do
not duplicate existing names. Returns a merged, de-duplicated list. \\
\texttt{finalize\_classify(prompt)} &
Labels the entire corpus against the current taxonomy and writes the
final \texttt{taxonomy.json} artifact. Gated by the
\texttt{min\_iterations} floor and refuses to re-run on an unchanged
taxonomy. \\
\bottomrule
\end{tabular}
\caption{The orchestrator's six-tool action space. All taxonomy
mutations flow through \texttt{revise\_taxonomy}; all per-item
labelling flows through the judge. \texttt{finalize\_classify} is the
only tool that writes the user-facing artifact.}
\label{tab:tools}
\end{table*}

Two design choices in this table are load-bearing. First,
\texttt{classify\_with\_judge} and \texttt{finalize\_classify} both
take a \emph{prompt string} written by the orchestrator at call time,
instead of a fixed system prompt. This lets the orchestrator
specialise the labelling instructions to the taxonomy it has built;
for instance, it can tell the judge to attend to rhetorical cues for a
discourse-style corpus, or to named entities for a topic-style
corpus, without us having to anticipate every domain. Second,
\texttt{propose\_novelties\_with\_judge} is a separate tool; novelty proposal is not a
flag on \texttt{classify\_with\_judge}, because the two prompt
shapes are different: classification is one-of-N with a free-form
escape, while novelty proposal is open-ended generation. Conflating
them hurt judge reliability in early prototypes.

\subsection{The Revise DSL}
\label{sec:system:revise}

Every taxonomy mutation flows through \texttt{revise\_taxonomy},
which accepts an ordered list of typed operations. The six
op shapes are:

\begin{itemize}
\setlength{\itemsep}{1pt}
\item \texttt{\{op: add, name, description\}}
\item \texttt{\{op: rename, old\_name, new\_name\}}
\item \texttt{\{op: edit, name, description\}}
\item \texttt{\{op: merge, into, from: [...], description?\}}
\item \texttt{\{op: split, from, into: [\{name, description\}, ...]\}}
\item \texttt{\{op: drop, name\}}
\end{itemize}

Each op is dispatched to a pure handler of signature
\texttt{(tax, op) $\to$ (new\_tax, log\_entry)}. The handlers are
registered in a single \texttt{\_OPS} table, which keeps ``add a new
op kind'' to ``write a function and register it''; the test suite
exercises each handler in isolation. Two invariants matter for
correctness:

\paragraph{Validate before mutate.} Each handler checks every
precondition before constructing a new taxonomy list. Two early bugs
motivated this: a \texttt{merge} that removed source categories before
discovering the target name was missing, and a \texttt{split} that
removed the source before discovering the \texttt{into} list was
malformed. Both lost data. The fix is invariant: if any precondition
fails, the handler returns the input taxonomy unchanged with a
descriptive log entry, and the orchestrator gets to read what went
wrong on the next step.

\paragraph{Per-op failure isolation.} The dispatcher records failures
per op and continues with the rest of the batch. A revise call of
``add A, rename B, merge X'' where the merge target is missing applies
the add and the rename and skips only the merge; the orchestrator
sees the log, decides whether the partial application is acceptable,
and either retries or moves on. This is much easier for the agent to
recover from than an all-or-nothing rejection, and the log makes the
trace human-auditable after the fact.

The DSL is small on purpose. Free-form taxonomy edits expressed in
prose would put parsing burden on us, hide intent from the trace, and
make it impossible to enforce the invariants above. The six verbs
themselves are the standard taxonomy-edit vocabulary
(cf.\ TaxoAdapt~\citep{taxoadapt2025}); the contribution here is the
typed dispatcher that validates each operation before applying it, with isolation between operations on failure,
not the verb set itself. Across the corpora in
\S\ref{sec:evaluation} we did not observe an orchestrator wanting an
op outside this set, but we have not run a controlled ablation
against edits expressed in free-form prose, and do not claim the six ops are
provably complete.

\subsection{Convergence}
\label{sec:system:convergence}

Convergence is decided by the judge.
After each
\texttt{classify\_with\_judge} call, the system computes the
\emph{don't-fit rate}, the fraction of items the judge labelled
\texttt{"other"}, and the orchestrator is allowed to call
\texttt{finalize\_classify} once that rate falls below a threshold
(default $0.10$). The $0.10$ default is a usability choice; it is not a
calibrated value; we report terminal don't-fit rates alongside
external purity and NMI in \S\ref{sec:evaluation} as a sanity check,
but do not claim the rate is a quantitatively validated leading
indicator. The threshold is also intentionally orchestrator-readable:
because the orchestrator authors the classify prompt, a permissive
prompt could in principle drive the rate down without genuine
convergence (we flag this as a limitation in
\S\ref{sec:limitations}). Three details make the signal usable in
practice.

\paragraph{Escape-hatch labels only.} Every classify
prompt is hardened with a fixed suffix instructing the judge to reply
\texttt{"other"} whenever no listed category fits, and to never invent
a new name. Judge replies are then mapped against the taxonomy
case-insensitively; any out-of-taxonomy label is coerced to
\texttt{"other"} and the rationale is annotated to flag the coercion.
This is what makes the don't-fit rate a meaningful convergence signal:
a high rate genuinely reflects items the taxonomy does not cover,
and does not reflect a judge that hallucinated novel labels.

\paragraph{Minimum-iteration floor.} A small probe can produce a
deceptively low don't-fit rate when the taxonomy is still half-built.
\texttt{finalize\_classify} therefore refuses to run until the
orchestrator has completed at least \texttt{min\_iterations}
classification rounds (default 3 in \texttt{run()}). This is a soft
constraint, since the orchestrator gets an error message and keeps
iterating, and it can be tuned per corpus.

\paragraph{Judge-error isolation.} If the judge HTTP call itself fails
after retries, the item is labelled \texttt{"other"} with the sentinel
rationale \texttt{[judge call failed]} and \emph{excluded} from the
denominator of the don't-fit rate. A transient API outage should not trick
the orchestrator into thinking the taxonomy stopped fitting. The
error count is reported back so the orchestrator (and the user) can
see what happened.

\subsection{Operational Safeguards}
\label{sec:system:safeguards}

Three engineering choices keep the system usable as a walk-up demo,
and remains usable as a tool beyond the research prototype phase.

\paragraph{Live cost tracking.} A \texttt{CostTracker} object receives
usage from every LLM call, including both the streamed orchestrator
\texttt{AIMessage}s and each judge HTTP reply, and writes
\texttt{cost.json} after every agent super-step. We pass OpenRouter's
\texttt{usage: \{include: true\}} flag so each reply carries the
\emph{actual} charge under \texttt{usage.cost}, and prefer that native
value over a static price table denominated per million tokens. The fallback table
covers the common defaults (Sonnet, Llama 3.3 70B, Gemini Flash, etc.)
for models that do not honour the flag.

\paragraph{Partial-progress saving.} The orchestrator can fail
mid-run for several reasons (recursion limit, API outage,
out-of-budget classify call), so the system writes everything it can
as it goes. \texttt{taxonomy\_state.json} is rewritten after every
\texttt{revise\_taxonomy} call with the current working taxonomy and
the classify-call counter. \texttt{classifications.jsonl} is streamed
row-by-row as judge replies arrive during \texttt{finalize\_classify},
under a write lock so concurrent appends from the thread pool stay
consistent. The consolidated \texttt{taxonomy.json} is written only
after \texttt{finalize\_classify} returns; if the process dies before
that, the user still keeps the latest taxonomy and whatever labels
arrived. The \texttt{trace.jsonl} log records every tool call, op
log, and judge batch in order, for after-the-fact inspection.

\paragraph{Streamlit UI.} A single-file Streamlit app
(\texttt{app.py}) wraps \texttt{run()} for non-CLI users. It exposes
the parameters that matter (orchestrator model, judge model,
\texttt{min\_iterations}, don't-fit threshold, probe size, target
taxonomy size hint), shows live cost and wall-time estimates above
the Start button, streams the working taxonomy as
\texttt{revise\_taxonomy} mutates it, and lists past runs from
\texttt{meta.json} for inspection. Importantly, the UI reads the same
on-disk artifacts the CLI writes, with no privileged
in-memory state, so a run started from the command line can be
inspected in the UI without re-running anything.
